---
title: "Week 1: R Programming Assignment - 50 points (LO1)"
author: "guanhua"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Use the following files and variable
- R Data Set: iris
- R Grouping Variable: Species
- R Numeric Variable: Sepal.Length
- R Numeric Variable #2 (used for Scatter Plot): Sepal.Width
- Warning: These will change from quarter to quarter.

# Step 1: Describe the Data
- List the structure of the data (str)
- Execute a summary of the data
- Print the first six records
```{r data-description}
str(iris)
summary(iris)
head(iris)
```

# Step 2: Box-Whisker Plots
- Plot a box plot of the numeric variable split by the grouping variable. The plot needs the following:
  - The MAIN TITLE of the box plot should be set to your name
  - Add a notch to the boxes
  - Add color to the boxes
```{r boxplot}
boxplot(Sepal.Length ~ Species,
        data = iris,
        notch = TRUE,
        col = c("lightblue", "lightgreen", "lightpink"),
        main = "guanhua")
```

# Step 3: Histograms
- Plot a histogram of the numeric variable
  - Manually set the number of breaks to a value that makes sense
  - Superimpose a density line to the graph
  - Push Yourself! Find a way to add color or decoration to this graph. Mention this work in your PDF/DOC file in order for the grader to see the work and give you credit.
```{r histogram}
hist(iris$Sepal.Length,
     breaks = 12,
     prob = TRUE,            # so the histogram can be compared to a density curve
     col = "lightblue",      # a bit of color decoration
     border = "black",       # adding borders for clarity
     main = "Histogram of Sepal Length",
     xlab = "Sepal Length")

# Superimpose a density curve (in red) onto the histogram
lines(density(iris$Sepal.Length), col = "red", lwd = 2)

# Optionally, add a rug plot for additional decoration
rug(iris$Sepal.Length, col = "darkblue")
```

# STEP 4: Scatter Plots
- Create a scatter plot of the numeric variable and variable #2.
  - Each group member needs to be in a different color
  - Set the plot character to a value that is easy to see. Character value 16 is a good choice.
  - Push Yourself! Find a way to add color or decoration to this graph. Mention this work in your PDF/DOC file in order for the grader to see the work and give you credit.
```{r scatterplot}
# Create a scatter plot for the two numeric variables:
# x-axis: Sepal.Length, y-axis: Sepal.Width
# Use a different color for each species and a common plotting character (pch = 16).
plot(iris$Sepal.Length, iris$Sepal.Width,
     col = as.numeric(iris$Species),   # different color for each species
     pch = 16,
     xlab = "Sepal Length",
     ylab = "Sepal Width",
     main = "Scatter Plot of Sepal Length vs. Sepal Width")

# Add a legend mapping colors to species
legend("topright",
       legend = levels(iris$Species),
       col = 1:length(levels(iris$Species)),
       pch = 16)

# Optionally, add a grid for enhanced decoration
grid()
```

# STEP 5: Simple Math
- For the numeric variable, compute the following statistics
  - Mean
  - Median
  - Min
  - Max
  - Standard Deviation
  - Calculate the Median for the numeric for each group member. Sort the result in Descending order.
```{r simple-math}
# Compute the basic statistics for the numeric variable Sepal.Length
sepal_length_mean <- mean(iris$Sepal.Length)
sepal_length_median <- median(iris$Sepal.Length)
sepal_length_min <- min(iris$Sepal.Length)
sepal_length_max <- max(iris$Sepal.Length)
sepal_length_sd <- sd(iris$Sepal.Length)

# Print the results
cat("Mean of Sepal Length:", sepal_length_mean, "\n")
cat("Median of Sepal Length:", sepal_length_median, "\n")
cat("Min of Sepal Length:", sepal_length_min, "\n")
cat("Max of Sepal Length:", sepal_length_max, "\n")
cat("Standard Deviation of Sepal Length:", sepal_length_sd, "\n\n")

# Compute the median of Sepal.Length for each species and sort in descending order
median_by_species <- aggregate(Sepal.Length ~ Species, data = iris, FUN = median)
median_by_species <- median_by_species[order(median_by_species$Sepal.Length, decreasing = TRUE), ]
print(median_by_species)
```
