---
title: "Week 2: R Programming Assignment - 50 points (LO2)"
author: "guanhua"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(width.cutoff = 60))
```

# Use the following file
- R Data Set: HMEQ_Loss.csv (in the zip file attached).
- The Data Dictionary in the zip file.
- Grouping Variable: TARGET_BAD_FLAG

# Step 1: Read in the Data
- Read the data into R
- List the structure of the data (str)
- Execute a summary of the data
- Print the first six records

```{r read_data}
# Read the data into R
hmeq_data <- read.csv("HMEQ_Loss.csv")

# List the structure of the data (str)
str(hmeq_data)

# Execute a summary of the data
summary(hmeq_data)

# Print the first six records
head(hmeq_data)
```

# Step 2: Box-Whisker Plots
- Plot a box plot of all the numeric variables split by the grouping variable. The plot needs the following:
  - The MAIN TITLE of the box plot should be set to your name
  - Add color to the boxes
  - Comment on whether or not there are any observable differences in the box plots between the two groups.

```{r box_plots}
# Identify numeric variables (excluding TARGET_BAD_FLAG as it's the grouping variable and TARGET_LOSS_AMT which is not used for grouping)
numeric_vars <- sapply(hmeq_data, is.numeric)
numeric_vars["TARGET_BAD_FLAG"] <- FALSE # Exclude TARGET_BAD_FLAG
numeric_vars["TARGET_LOSS_AMT"] <- FALSE # Exclude TARGET_LOSS_AMT
numeric_col_names <- names(hmeq_data)[numeric_vars]

# Create box plots for each numeric variable split by TARGET_BAD_FLAG
par(mfrow=c(3, 4)) # Adjust layout as needed to fit all boxplots

for (col_name in numeric_col_names) {
  boxplot(hmeq_data[[col_name]] ~ hmeq_data$TARGET_BAD_FLAG,
          main = paste(col_name, "by guanhua"),
          xlab = "TARGET_BAD_FLAG",
          ylab = col_name,
          col = c("lightblue", "lightgreen")) # Add color to boxes
}

par(mfrow=c(1, 1)) # Reset to default layout

# Comment on observable differences:
# From the box plots, we can observe some potential differences between the two groups (TARGET_BAD_FLAG = 0 and TARGET_BAD_FLAG = 1)
# for several numeric variables. For example, LOAN, MORTDUE, VALUE, DEROG, DELINQ, NINQ, and DEBTINC seem to have different distributions
# between the two groups. Specifically, it appears that the group with TARGET_BAD_FLAG = 1 (bad loans) tends to have:
# - Lower LOAN amounts
# - Lower VALUE of the house
# - Higher DEROG and DELINQ counts
# - Slightly higher NINQ (number of inquiries)
# - Higher DEBTINC (debt to income ratio)
# These are just visual observations and would need further statistical testing to confirm significance.
# Variables like YOJ and CLAGE might also show some differences but are less pronounced.
```

# Step 3: Histograms
- Plot a histogram of at least one of the numeric variables
  - Manually set the number of breaks to a value that makes sense
  - Superimpose a density line to the graph

```{r histograms}
# Plot a histogram of the LOAN variable
hist(hmeq_data$LOAN,
     breaks = 20,
     main = "Histogram of LOAN",
     xlab = "Loan Amount",
     ylab = "Frequency",
     col = "skyblue",
     freq = TRUE) # Ensure frequency is plotted

# Superimpose a density line to the graph
lines(density(hmeq_data$LOAN, na.rm = TRUE),
      col = "red",
      lwd = 2)
```

# Step 4: Impute "Fix" all the numeric variables that have missing values
- For the missing Target variables, simply set the missing values to zero
- For the remaining numeric variables with missing values, create two new variables. One variable will have a name beginning with IMP_ and it will contained the imputed value. The second value will have a name beginning with M_ and it will contain a 1 if the record was imputed and a zero if it was not.
- You may impute with any method that makes sense. The median or mean value will be useful in most cases.
- Push yourself! Try one complex imputation like the one described in the lectures.
- Delete the original variable after it has been imputed.
- Run a summary to prove that all the variables have been imputed
- Compute a sum for all the M_ variables to prove that the number of flags is equal to the number of missing values.

```{r impute_missing_values}
# --- Impute missing values ---

# Fix missing Target variables (TARGET_LOSS_AMT) to zero
hmeq_data$TARGET_LOSS_AMT[is.na(hmeq_data$TARGET_LOSS_AMT)] <- 0

# Identify numeric variables with missing values (excluding Target variables as they are handled separately)
numeric_vars_missing <- numeric_col_names[colSums(is.na(hmeq_data[, numeric_col_names])) > 0]

# Impute missing values for each numeric variable
for (col_name in numeric_vars_missing) {
  # --- Simple Imputation (Median) ---
  median_val <- median(hmeq_data[[col_name]], na.rm = TRUE)
  hmeq_data[[paste0("IMP_", col_name)]] <- ifelse(is.na(hmeq_data[[col_name]]), median_val, hmeq_data[[col_name]])

  # --- Missing Value Flag ---
  hmeq_data[[paste0("M_", col_name)]] <- ifelse(is.na(hmeq_data[[col_name]]), 1, 0)

  # --- Complex Imputation Mean imputation based on JOB ---
  mean_by_job <- aggregate(hmeq_data[[col_name]] ~ hmeq_data$JOB, data = hmeq_data, FUN = function(x) mean(x, na.rm = TRUE))
  names(mean_by_job) <- c("JOB", paste0("MEAN_", col_name))
  hmeq_data <- merge(hmeq_data, mean_by_job, by = "JOB", all.x = TRUE)
  hmeq_data[[paste0("IMP_", col_name)]] <- ifelse(is.na(hmeq_data[[col_name]]),
                                                  hmeq_data[[paste0("MEAN_", col_name)]],
                                                  hmeq_data[[col_name]])
  hmeq_data <- hmeq_data[, !(names(hmeq_data) %in% paste0("MEAN_", col_name))] # Remove MEAN_ variable after imputation

  # --- Delete Original Variable ---
  hmeq_data[[col_name]] <- NULL
}

# --- Run a summary to prove that all the variables have been imputed ---
summary(hmeq_data)

# --- Compute a sum for all the M_ variables to prove that the number of flags is equal to the number of missing values ---
m_vars_names <- names(hmeq_data)[startsWith(names(hmeq_data), "M_")]
missing_counts <- colSums(hmeq_data[, m_vars_names])
print("Sum of M_ variables (Missing Value Counts):")
print(missing_counts)

# Verify against original missing values count (optional - for verification)
original_missing_counts <- colSums(is.na(read.csv("HMEQ_Loss.csv")[,numeric_col_names]))
print("Original Missing Value Counts:")
print(original_missing_counts[names(original_missing_counts) %in% numeric_vars_missing])
```


# Step 5: One Hot Encoding
- For the character / category variables, perform one hot encoding. For this create a Flag for each categories.
- Delete the original class variable
- Run a summary to show that the category variables have been replaced by Flag variables.

```{r one_hot_encoding}
# Identify character/category variables
categorical_vars <- names(hmeq_data)[sapply(hmeq_data, is.character) | sapply(hmeq_data, is.factor)]

# Perform one-hot encoding for each categorical variable
for (col_name in categorical_vars) {
  # Create dummy variables using model.matrix
  dummies <- model.matrix(~ . -1, data = data.frame(Category = hmeq_data[[col_name]]))
  dummies_df <- as.data.frame(dummies)
  names(dummies_df) <- paste0("FLAG_", col_name, "_", colnames(dummies_df)) # Rename columns

  # Combine with the main dataframe
  hmeq_data <- cbind(hmeq_data, dummies_df)

  # Delete the original categorical variable
  hmeq_data[[col_name]] <- NULL
}

# --- Run a summary to show that the category variables have been replaced by Flag variables. ---
summary(hmeq_data)
